An Effective Model of HL-60 Differentiation

### 2017-01-24 Paper_v2.tex
updating the MM

\subsubsection*{Effective ATRA differentiation model.}
ATRA induced signaling events were modeled using saturation kinetics within an ordinary differential equation (ODE) framework:
\begin{equation}
	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\left(\mu + k_{d}\right) x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}
\end{equation}
The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model, respectively.
The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of process $j$.
Typically, process $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
Lastly, $\mu$ denotes the specific growth rate, and $k_{d}$ denotes the rate constant controlling cell death.
Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.

Signaling rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the HL-60 model.
The rate of an enzyme catalyzed process was modeled using saturation kinetics:
\begin{equation}\label{eqn:rate-saturation}
	\bar{r}_{j} = k_{j}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\left(\frac{x_{s}}{K_{js} + x_{s}}\right)
\end{equation}
where $k_{j}$ denotes the catalytic rate constant for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
and $K_{js}$ denotes the saturation constant for species $s$ and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$.
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
If a process had no modifying factors, $v_{j}=1$.

@@

The HL-60 model described both signal transduction and gene expression events following the addition of ATRA.
The output of the signal transduction model was the input to the gene expression model.
For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$):
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
\end{eqnarray}
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
The specific transcription rate was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic rate of transcription was modeled as:
\begin{equation}
	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$. The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths,
where $L_{T,o}$ is a characteristic gene length, and $L_{T,j}$ denotes the length of gene $j$.
Thus, the $\left(L_{T,o}/L_{T,j}\right)$ term is gene specific correction to the characteristic transcription rate.
The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.

The gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$,
and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule.
In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations resulting in gene expression.
Thus, the control variable $u_{j}$ took the form:
\begin{equation}
	{u}_{j} = \frac{W_{R_{1},j}+\displaystyle\sum_{n}W_{nj}f_{nj}}{1+W_{R_{1},j}+\displaystyle\sum_{d}W_{dj}f_{dj}}
\end{equation}where the numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using hill like functions), while
the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
If a gene expression process had no modifying factors, $u_{j}=1$.
Lastly, the specific translation rate was modeled as:
\begin{equation}
	r_{X,j} = V^{max}_{X}\left(\frac{m_{j}}{K_{X}+m_{j}}\right)
\end{equation}
where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, and $K_{X}$ denotes a translation saturation constant.
The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
and the Ribosome abundance ($R_{2}$), $V^{max}_{X} = k_{X}\left(R_{2}\right)$.

In this study, we estimated the $W_{ij}$ parameters, and the parameters in the transfer functions $f_{dj}$ from gene expression data sets.
On the other hand, we estimated $k_{T}, k_{X}, \theta_{m,j}, \theta_{p,j}, R_{1}$ and $R_{2}$ using estimates of transcription and translation rates,
the half-life of a typical mRNA and protein, and a typical value for the copies per cell of RNA polymerase and ribosomes from literature \cite{Milo:2010aa}.
The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
The values used for the characteristic transcription/translation parameters, degradation constants and
macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.

\noindent\subsubsection*{Estimation of signaling and gene expression model parameters.}
Signal and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental data set $j$:
\begin{equation}\label{eqn:objective-function}
	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
\end{equation}
The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs at time $i$ from training set $j$,
where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
The first term in Eqn. \eqref{eqn:objective-function} quantified the relative simulation error.
We used immunoblot intensity measurements for model training.
Thus, we trained the model on the \emph{relative} change between bands within each data set.
Suppose we have the intensity of species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$.
The scaled value $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ is given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
\end{equation}
where $\hat{\mathcal{M}}_{ij}=0$ and $\hat{\mathcal{M}}_{ij}=1$  describe the lowest (highest) intensity bands.
A similar scaling was used for the simulation output. The second term in the objective function ensured a realistic concentration scale was estimated by the model.
We set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
We minimized the total model residual $\sum_{j}E_{j}$ using heuristic optimization starting from a random initial parameter guess.

The signaling and gene expression model equations were implemented in Julia and solved using the \texttt{CVODE} routine of the Sundials package \cite{Julia,Hindmarsh2005}.
The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.

>>>>

\subsubsection*{Model equations.}
We decomposed the ATRA-induced differentiation program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RXR/RAR (activated Trigger) signals;
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
The output of the signal initiation module was the input to the gene expression model.
For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$) and protein ($p_{j}$) abundance:
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
\end{eqnarray}
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
The specific transcription rate $r_{T,j}$ was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
\begin{equation}
	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
The degradation rate constants were defined as $\theta_{m,j}$ and $\theta_{p,j}$ denote characteristic degradation constants for mRNA and protein, respectively.
Lastly, the $\lambda_{j}$ term denotes the constitutive rate of expression of gene $j$.

The gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced the expression of gene $j$.
If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$,
and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule which combines the individual regulatory inputs for gene $j$ into a single control term.
In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations that resulted in gene expression [REFHERE]:
\begin{equation}
	{u}_{j} = \frac{W_{R_{1},j}+\displaystyle\sum_{n}W_{nj}f_{nj}}{1+W_{R_{1},j}+\displaystyle\sum_{d}W_{dj}f_{dj}}
\end{equation}
The numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using hill like functions), while
the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
Given this formulation, the rate of constitutive expression was then given by:
\begin{equation}
	\lambda_{j} = \bar{r}_{T,j}\left(\frac{W_{R_{1},j}}{1+W_{R_{1},j}}\right)
\end{equation}
If a gene expression process had no modifying factors, $u_{j}=1$.
Lastly, the specific translation rate was modeled as:
\begin{equation}
	r_{X,j} = V^{max}_{X}\left(\frac{L_{X,o}}{L_{X,j}}\right)\left(\frac{m_{j}}{K_{X}+m_{j}}\right)
\end{equation}
where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, and $K_{X}$ denotes a translation saturation constant.
The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
and the Ribosome abundance ($R_{2}$), $V^{max}_{X} = k_{X}\left(R_{2}\right)$. As was the case for transcription,
we corrected the characteristic translation rate by the ratio of the length of a characteristic transcription normalized by the length of transcript $j$.

The signal initiation, and integration modules required the level of cRaf-pS621 and ATRA-RXR/RAR (activated Trigger) as inputs.
However, our base model described only the abundance of inactive proteins e.g., cRaf or RXR/RAR but not the activated forms.
To address this issue, we estimated pseudo steady state approximations for the abundance of cRaf-pS621 and activated Trigger.
The abundance of activated trigger ($x_{a,1}$) was estimated directly from the RXR/RAR abundance ($x_{u,1}$):
\begin{equation}
	x_{a,1} \sim x_{u,1}\left(\frac{\alpha\times{\texttt{ATRA}}}{1+\alpha\times{ATRA}}\right)
\end{equation}
On the other hand, the abundance of cRaf-pS621 was estimated by making the pseudo steady state approximation on the cRaf-pS621 balance.
The abundance of an activated signaling species $i$ was given by:
\begin{equation}
	\frac{dx_{i}}{dt}  =  r_{+,i}\left(\mathbf{x},\mathbf{k}\right) - \left(\mu+k_{d,i}\right)x_{i}\qquad{i=1,\hdots,\mathcal{M}}\\
\end{equation}
The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model,
respectively. The term $r_{+,i}\left(\mathbf{x},\mathbf{k}\right)$ denotes the rate of generation of activated species $i$,
while  $\mu$ denotes the specific growth rate, and $k_{d,i}$ denotes the rate constant controlling the non-specific degradation of $x_{i}$.
We have neglected deactivation reactions e.g., phosphatase activities.
We assumed that signaling processes were fast compared to gene expression; this allowed us to approximate the signaling balance as:
\begin{equation}
	x_{i}^{*} \simeq \frac{r_{+,i}\left(\mathbf{x},\mathbf{k}\right)}{\left(\mu+k_{d,i}\right)} \qquad{i=1,\hdots,\mathcal{M}}
\end{equation}
The generation rate was written as the product of a kinetic term ($\bar{r}_{+,i}$) and a control term ($v_{i}$).
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
If a process had no modifying factors, $v_{j}=1$.
The kinetic rate of cRaf-pS621 generation $\bar{r}_{+,cRaf}$ was given as:
\begin{equation}
	\bar{r}_{+,cRaf} = k_{+,cRaf}x_{s}\left(\frac{x_{cRaf}}{K_{+,cRaf}+x_{cRaf}}\right)
\end{equation}where $x_{s}$ denotes the signalsome abundance, and $K_{+,cRaf}$ denotes a saturation constant governing cRaf-pS621 formation.
The formation of cRaf-pS621 was regulated by a single factor, the abundance of MAPK inhibitor, thus $v_{+,cRaf}$ took the form:
\begin{equation}
	v_{+,cRaf} = \left(1-\frac{I}{K_{D}+I}\right)
\end{equation}where $I$ denotes the MAPK inhibitor abundance, and $K_{D}$ the affinity constant for inhibitor $I$.

% ATRA induced signaling events were modeled using saturation kinetics within an ordinary differential equation (ODE) framework:
% \begin{equation}
% 	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\left(\mu + k_{d}\right) x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}
% \end{equation}
% The quantity $x_{i}$ denotes concentration of signaling species $i$, while $\mathcal{R}$ and $\mathcal{M}$ denote the number of signaling reactions and signaling species in the model, respectively.
% The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
% The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of process $j$.
% Typically, process $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
% The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
% If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
% Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
% Lastly, $\mu$ denotes the specific growth rate, and $k_{d}$ denotes the rate constant controlling cell death.
% Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
%
% Signaling rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the HL-60 model.
% The rate of an enzyme catalyzed process was modeled using saturation kinetics:
% \begin{equation}\label{eqn:rate-saturation}
% 	\bar{r}_{j} = k_{j}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\left(\frac{x_{s}}{K_{js} + x_{s}}\right)
% \end{equation}
% where $k_{j}$ denotes the catalytic rate constant for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
% and $K_{js}$ denotes the saturation constant for species $s$ and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$.
% The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
% For each rate, we used a rule-based approach to select from competing control factors.
% If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
% $v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
% where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
% The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
% variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010138}.
% If a process had no modifying factors, $v_{j}=1$.

@@

% The HL-60 model described both signal transduction and gene expression events following the addition of ATRA.

\noindent\subsubsection*{Estimation of signaling and gene expression model parameters.}
We estimated the $W_{ij}$ parameters, and the parameters in the transfer functions $f_{dj}$ from gene expression data sets.
On the other hand, we estimated $k_{T}, k_{X}, \theta_{m,j}, \theta_{p,j}, R_{1}$ and $R_{2}$ using estimates of transcription and translation rates,
the half-life of a typical mRNA and protein, and a typical value for the copies per cell of RNA polymerase and ribosomes from literature \cite{Milo:2010aa}.
The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
Lastly, we calculated the concentration for gene $G_{j}$ by assuming, on average, that a cell had two copies of each gene at any given time.
Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
The values used for the characteristic transcription/translation parameters, degradation constants and
macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.

Signal and gene expression model parameters were estimated by minimizing the squared difference between simulations and experimental data set $j$:
\begin{equation}\label{eqn:objective-function}
	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2 + \left(\frac{{\mathcal{M}^{\prime}_{ij}}-\max{y_{ij}}}{{\mathcal{M}^{\prime}_{ij}}}\right)^{2}
\end{equation}
The terms $\hat{\mathcal{M}}_{ij}$ and $\hat{y}_{ij}$ denote scaled experimental observations and simulation outputs at time $i$ from training set $j$,
where $\mathcal{T}_{j}$ denoted the number of time points for data set $j$.
The first term in Eqn. \eqref{eqn:objective-function} quantified the relative simulation error.
We used immunoblot intensity measurements for model training.
Thus, we trained the model on the \emph{relative} change between bands within each data set.
Suppose we have the intensity of species $x$ at time $\{t_{1},t_{2},..,t_{n}\}$ in condition $j$.
The scaled value $0\leq\hat{\mathcal{M}}_{ij}\leq{1}$ is given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \left({\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}\right)/\left({\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}\right)
\end{equation}
where $\hat{\mathcal{M}}_{ij}=0$ and $\hat{\mathcal{M}}_{ij}=1$  describe the lowest (highest) intensity bands.
A similar scaling was used for the simulation output. The second term in the objective function ensured a realistic concentration scale was estimated by the model.
We set the highest intensity band to $\mathcal{M}^{\prime}_{ij} = 10$ [AU] for all simulations.
We minimized the total model residual $\sum_{j}E_{j}$ using heuristic optimization starting from a random initial parameter guess.

The signaling and gene expression model equations were implemented in the Julia programming language, and solved using the \texttt{CVODE} routine of the Sundials package \cite{Julia,Hindmarsh2005}.
The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
